---
title: "Introduction to the L2E Package (version 2.0)"
author: "Xiaoqian Liu, Jocelyn T. Chi, Lisa Lin, Kennth Lange, and Eric C. Chi"
bibliography: l2e-ref.bib
output:
  pdf_document: default
  html_document: null
  toc: yes
  toc_float:
    collapsed: no
biblio-style: unsrt
vignette: |
  %\VignetteIndexEntry{Introduction to the L2E Package} 
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

<style>
img{ background:none; border:none; box-shadow:none; }
</style>

\[
\newcommand{\vbeta}{\boldsymbol{\beta}}
\]

```{r echo=FALSE, cache=FALSE}
set.seed(12345)
knitr::opts_chunk$set(
  cache=TRUE,
  comment = '', 
  fig.width = 5, 
  fig.height = 5,
  fig.align='center'
)
```
## Overview

The `L2E` package (version 2.0) implements the computational framework for L$_2$E regression in @L2E-Liu, which was built on the previous work in @L2E-Chi. Both work use the block coordinate descent  strategy to solve a nonconvex optimization problem but employ different methods for the inner block descent updates. We refer to the method in @L2E-Liu as "MM" and the one in @L2E-Chi as "PG" in our package. This vignette provides code to replicate some examples illustrating the usage of the frameworks in both papers and the improvements from @L2E-Liu.

## Examples

### Isotonic Regression

The first example illustrates how to perform robust isotonic regression via the L$_{2}$ criterion, also called L$_2$E isotonic regression.  We begin by generating the true fit $f$ from a cubic function (depicted by the black line on the figure below).  We obtain the observed response by adding some Gaussian noise to $f$ (depicted by the gray points on the figure below).

```{r}
set.seed(12345)
n <- 200
tau <- 1
x <- seq(-2.5, 2.5, length.out=n)
f <- x^3
y <- f + (1/tau)*rnorm(n)
plot(x, y, pch=16, col='gray')
lines(x, f, lwd=3)
```

The `L2E_isotonic` function provides two options for implementing the L$_2$E isotonic regression. By setting the argument "method = MM" (the default), the function calls `l2e_regression_isotonic_MM` to perform L$_2$E isotonic regression using the MM method in @L2E-Liu. Setting "method = PG" leads the function to call `l2e_regression_isotonic` to perform L$_2$E isotonic regression using the PG method in @L2E-Chi. 

We obtain and compare three estimates for the underlying fit $f$ from the classical least squares (LS), the L$_{2}$E with MM,  and the L$_{2}$E with PG, respectively.  The true fit is shown in black while the fits from the LS, the L$_{2}$E with MM, and the  L$_{2}$E with PG are shown in blue, red, and green, respectively.  In the absence of outliers, all methods produce similar estimates. For the two L$_{2}$E methods, MM is faster than PG.

```{r}
library(L2E)
library(isotone)
tau <- 1/mad(y)
b <- y
# LS method
iso <- gpava(1:n, y)$x
# MM method
sol_mm <- L2E_isotonic(y, b, tau, method = "MM")
# PG method
sol_pg <- L2E_isotonic(y, b, tau, method = 'PG')


# Plots
plot(x, y, pch=16, col='gray')
lines(x, f, lwd=3)
lines(x, iso, col='blue', lwd=3) ## LS
lines(x, sol_mm$beta, col='red', lwd=3) ## MM
lines(x, sol_pg$beta, col='green', lwd=3) ## PG
legend("bottomright", legend = c("LS", "MM", "PG"), col = c('blue','red', 'green'), lwd=3)
```

Next, we introduce some outliers by perturbing some of the observed responses.

```{r}
num <- 20
ix <- 1:num
y[45 + ix] <- 14 + rnorm(num)
plot(x, y, pch=16, col='gray')
lines(x, f, lwd=3)
```

We obtain the three estimates with the outlying points.  The L$_{2}$E fits are less sensitive to outliers than the LS fit in this example. The two methods MM and PG produce comparable estimates, but MM is faster than PG.

```{r}

tau <- 1/mad(y)
b <- y
# LS method
iso <- gpava(1:n, y)$x
# MM method
sol_mm <- L2E_isotonic(y, b, tau, method = "MM")
# PG method
sol_pg <- L2E_isotonic(y, b, tau, method = 'PG')


# Plots
plot(x, y, pch=16,  col='gray')
lines(x, f, lwd=3)
lines(x, iso, col='blue', lwd=3) ## LS
lines(x, sol_mm$beta, col='red', lwd=3) ## MM
lines(x, sol_pg$beta, col='green', lwd=3) ## PG
legend("bottomright", legend = c("LS", "MM", "PG"), col = c('blue','red', 'green'), lwd=3)
```


### Convex Regression

The second example illustrates how to perform robust convex regression via the L$_{2}$ criterion, also called L$_2$E convex regression.  We begin by simulating a convex function $f$.  The observed response is composed of $f$ with some additive Gaussian noise.

```{r}
set.seed(12345)
n <- 300
tau <- 1
x <- seq(-2, 2, length.out=n)
f <- x^4 + x
y <- f + (1/tau) * rnorm(n)
plot(x, y, pch=16, col='gray', cex=0.8)
lines(x, f, col='black', lwd=3)
```

The `L2E_convex` function provides two options for implementing the L$_2$E convex regression. By setting the argument "method = MM" (the default), the function calls `l2e_regression_convex_MM` to perform L$_2$E convex regression using the MM method in @L2E-Liu. Setting "method = PG" leads the function to call `l2e_regression_convex` to perform L$_2$E convex regression using the PG method in @L2E-Chi. 

We obtain and compare three estimates for the underlying fit $f$ from the classical least squares (LS), the L$_{2}$E with MM,  and the L$_{2}$E with PG, respectively. The true fit is shown in black while the fits from the LS, the L$_{2}$E with MM,  and the  L$_{2}$E with PG are shown in blue, red, and green, respectively.  In the absence of outliers, all methods produce similar estimates. For the two  L$_{2}$E methods, MM is faster than PG.

```{r}
library(cobs)
tau <- 1/mad(y)
b <- y
## LS method
cvx <- fitted(conreg(y, convex=TRUE))
## MM method
sol_mm <- L2E_convex(y, b, tau, method = "MM")
## PG method
sol_pg <- L2E_convex(y, b, tau, method = 'PG')

plot(x, y, pch=16, col='gray')
lines(x, f, lwd=3)
lines(x, cvx, col='blue', lwd=3) ## LS
lines(x, sol_mm$beta, col='red', lwd=3) ## MM
lines(x, sol_pg$beta, col='green', lwd=3) ## PG
legend("bottomright", legend = c("LS", "MM", "PG"), col = c('blue','red', 'green'), lwd=3)
```

Next, we again introduce some outlying points.

```{r}
num <- 50
ix <- 1:num
y[45 + ix] <- 14 + rnorm(num)
 
plot(x, y, pch=16, col='gray', cex=0.8)
lines(x, f, col='black', lwd=3)
```

We obtain the three estimates with the outlying points.  The L$_{2}$E fits are less sensitive to outliers than the LS fit in this example. The MM method not only requires less computional time but also produces better estimation than the PG method.

```{r}
tau <- 1/mad(y)
b <- y
## LS method
cvx <- fitted(conreg(y, convex=TRUE))
## MM method
sol_mm <- L2E_convex(y, b, tau, method = "MM")
## PG method
sol_pg <- L2E_convex(y, b, tau, method = 'PG')

plot(x, y, pch=16, col='gray')
lines(x, f, lwd=3)
lines(x, cvx, col='blue', lwd=3) ## LS
lines(x, sol_mm$beta, col='red', lwd=3) ## MM
lines(x, sol_pg$beta, col='green', lwd=3) ## PG
legend("bottomright", legend = c("LS", "MM", "PG"), col = c('blue','red', 'green'), lwd=3)

```



### Multivariate Regression
```{r echo=FALSE, cache=FALSE}
set.seed(12345)
knitr::opts_chunk$set(
  cache=TRUE,
  comment = '', 
  fig.width = 5, 
  fig.height = 4,
  fig.align='center'
)
```
The third example provides code to perform L$_{2}$E multivariate regression with the Hertzsprung-Russell diagram data of star cluster CYG OB1. The data set is commonly used in robust regression owing to its four known outliers -- four bright giant stars observed at low temperatures. The data set is publicly available in the R package `robustbase`. We begin by loading the data from the `robustbase` package.

```{r, title="star"}
library(robustbase)
data(starsCYG)
plot(starsCYG)
```

We use the `L2E_multivariate` function to perform the L$_2$E multivariate reeegression, namely, multivariate regression via the L$_{2}$ criterion.
The `L2E_multivariate` function provides two options for implementing the L$_2$E multivariate regression. By setting the argument "method = MM" (the default), the function calls `l2e_regression_MM` to perform L$_2$E multivariate regression using the MM method in @L2E-Liu. Setting "method = PG" leads the function to call `l2e_regression` to perform L$_2$E multivariate regression using the PG method in @L2E-Chi. 

We next use the MM method to fit the L$_2$E multivariate regression model and compare it to the least squares fit. The figure below shows that the L$_2$E fit can reduce the influence of the four outliers and fits the remaining data points well. The LS fit is heavily affected by the four outliers.

```{r}
y <- starsCYG[, "log.light"]
x <- starsCYG[, "log.Te"]
X0 <- cbind(rep(1, length(y)), x)

# LS method
mle <- lm(log.light ~ log.Te, data = starsCYG)
r_lm <- y - X0 %*% mle$coefficients

# L2E+MM method
tau <- 1/mad(y)
b <- c(0, 0)
# Fit the regression model
sol_mm <- L2E_multivariate(y, X0, b, tau, method="MM")
l2e_fit_mm <- X0 %*% sol_mm$beta
# compute limit weights
r_mm <- y - l2e_fit_mm


data <- data.frame(x, y, l2e_fit_mm)


d_lines <- data.frame(int = c(sol_mm$beta[1], mle$coefficients[1]),
                      sl = c(sol_mm$beta[2], mle$coefficients[2]),
                      col = c("red", "blue"),
                      lty = c("solid", "dashed"),
                      method = c("L2E", "LS"))

ltys <- as.character(d_lines$lty)
names(ltys) <- as.character(d_lines$lty)

cols <- as.character(d_lines$col)
cols <- cols[order(as.character(d_lines$lty))]
method <-  as.character(d_lines$method)


library(ggplot2)
library(latex2exp)
p <- ggplot() +
  geom_point(data = data, aes(x, y), size=2.5) + ylim(2, 6.5)+
  geom_abline(data = d_lines[d_lines$col == "red", ], 
              aes(intercept = int, slope = sl, lty = lty), color = "red", size=1) +
  geom_abline(data = d_lines[d_lines$col == "blue", ], 
              aes(intercept = int, slope = sl, lty = lty), color = "blue", size=1) +
  scale_linetype_manual(name = "Method", values = ltys, breaks=c("dashed", "solid"),
                        labels = c("LS  ", 
                                   expression(L[2]~E)),
                        guide = guide_legend(override.aes = list(colour = cols), legend=method))+
  theme_bw()

print(p)
```

We next show how to use the convereged weights from the MM method to identify outliers in the data. As discussed in @L2E-Liu, a small weight suggests a potential outlier.

```{r}
w <- as.vector(exp(-0.5* (sol_mm$tau*r_mm)**2 ))
data <- data.frame(x, y, l2e_fit_mm, w)
ggplot(data, aes(x=log10(w))) + geom_histogram()+
  labs(
    y="Count", x=expression(log[10]~'(w)'))+theme_bw()
```

The four extremely small weights in the above histogram indicate that there are four potential outliers in the data. Next, we incorporate the four outliers into the scatter plot. The MM method successfully identify the four outliers.

```{r}
outlier_mm <- rep("yes", length(y))
for (k in 1:length(y)) {
  if(w[k]>1e-5) # the threshold value can range from 1e-3 to 1e-14 according to the histogram
    outlier_mm[k] <- "no"
}
outlier_mm  <- factor(outlier_mm , levels=c("yes", "no"))
data <- data.frame(x, y, l2e_fit_mm, outlier_mm)


p+
  geom_point(data = data, aes(x, y, color=outlier_mm), size=2.5) +
  scale_color_manual(values = c(2,1), name="Outlier")+
  labs(
    y="Light Intensity", x="Temperature")+ theme_bw() 

```



### Other Examples

The new `L2E` package include implementation of L$_2$E sparse regression and L$_2$E trend filtering.
The `L2E_sparse_ncv` function  computes a solution path of L$_2$E sparse regression with exisitng penalties available in the `ncvreg` package. The  `L2E_sparse_dist` function computes a solution path of L$_2$E sparse regression with the distance penalty. The `L2E_TF_lasso` function computes a solution path of L$_2$E trend filtering with the Lasso penalty.  The `L2E_TF_dist` function computes a solution path of L$_2$E trend filtering with the distance penalty.  Readers may refer to @L2E-Liu for advantages of distance penalization in the two examples of sparse regression and trend filtering.





## References
